---
title: " Vignette illustrating the use of MOFA on the CLL data"
author: "Britta Velten"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Example_CLL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
devtools::load_all("/Users/bvelten/Documents/MOFA/MOFApackage/scGFA/MOFAtools/")
# devtools::load_all("/Users/ricard/mofa/MOFAtools")
library(MultiAssayExperiment)
```

# Step 1: Initialize MOFA object

## Load data
There are two ways to load the data into MOFA: 
(1) A list of matrices where features are rows and samples are columns: MutliAssayExperiment is constructed internally
(2) A MultiAssayExperiment object.
I WOULD INCLUDE HERE THE FILTERING MINVIEWS. 
```{r}
data("CLL_views")

# Remove patients from experiments in which they are completly unobserved
CLL_omics <-  lapply( CLL_views[c("lincRNA","meth","miRNA","mRNA","mut","viab")], 
                      function(mat) mat[,apply(mat,2, function(p) !all(is.na(p)))]) 

# Define pData
covariates <- data.frame(
  sex = CLL_views[["covariates"]][1,], 
  diagnosis = CLL_views[["covariates"]][2,]
)

# Define sampleMap
# (...)

# Create MultiAssayExperiment object
mae_CLL <- MultiAssayExperiment(experiments = CLL_omics, pData = covariates)
```

## Create MOFA object
The input data is stored in the InputData slot of the MOFA object.
```{r}
object <- createMOFAobject(mae_CLL)
# object <- createMOFAobjectFromList(ExpList = CLL_omics, pData = covariates)
```

# Step 2: Fit the MOFA model

## Define options

### Define I/O options
Important:
- tmpDir: Temporary folder to store the input matrices
- outFile: Output file for the model
- mofaDir: directory of the MOFA package
```{r}
DirOptions <- list(
  "tmpDir" = "/tmp",
  "outFile" = "/Users/ricard/test.hdf5",
  "mofaDir" = "/Users/ricard/mofa/MOFA"
)

DirOptions <- list(
  "tmpDir" = "/users/bvelten/Documents/MOFA/CLL_MOFA_data/testPipeline/",
  "outFile" = "/users/bvelten/Documents/MOFA/CLL_MOFA_data/testPipeline/",
  "mofaDir" = "/users/bvelten/Documents/MOFA/MOFApackage/scGFA/MOFA"
)
```

### Define model options
Important:
- Initial number of latent variables
- Likelihoods
- Schedule of updates
```{r}
ModelOptions <- getDefaultModelOpts(object)
ModelOptions$initialK <- 10
```

### Define training options
Important:
- Number of trials
- Maximum number of iterations
- Dropping options of latent factors
```{r}
TrainOptions <- getDefaultTrainOpts()
TrainOptions$maxiter <- 5
TrainOptions$drop_by_norm <- 0.00
TrainOptions$drop_by_r2 <- 0.00
```

## Prepare MOFA
Internally, prepareMOFA performs a set of sanity checks, fills the TrainOpts and ModelOpts slots of the MOFAmodel object and it also creates a set of temporary .txt files with the input matrices that will be loaded by the Python core implementation
```{r}
object <- prepareMOFA(object, 
                      DirOptions = DirOptions,
                      ModelOptions = ModelOptions,
                      TrainOptions = TrainOptions
)
```

## Run MOFA
This step can take some time , it is recommended to run it from the Python-based module
```{r}
object <- runMOFA(object, DirOptions)
```

# Step 3: Analyse a trained MOFA model

## Show the variance explained for each factor in each view
```{r}
calculateVarianceExplained(object)
```

## Scatterplots of pairs of latent variables
```{r}
# scatterPairs(object)
sctterPlot(object, facotr)
```

## Characterisation of specific latent factors

### Show weights and identify most relevant features
```{r}
showWeightHeatmap(object, "view")
showWeights(object, "view", factor, ntop, ntail, manual)
```

### Show original data for the genes with top weights
```{r}
view <- "mRNA"
factor <- "2"
showDataHeatmap(model, view, factor, nfeatures=50)

# The X axis is the latent variable and the Y axis is the observed value of the feature
# showDataScatter(model, view, factor, nfeatures=50, colour_by=NULL, shape_by=NULL)
```


### Feature set enrichment analysis
The characterisation of factors usually requires the inspection of the corresponding weights to identify features with strong influence. However, looking at single features can be challenging and often the combination of signal from functionally related sets of features (i.e. gene ontologies) is required.
Here we implemented a feature set enrichment analysis method (FeatureSetEnrichmentAnalysis) derived from the PCGSE package (REF).
The input of these function is a MOFA trained model (MOFAmodel), the factors for which to perform feature set enrichment (character vector), the feature sets (matrix) and a set of options regarding how the analysis should be performed. For more information see ...

Here we are going to perform feature set enrichment analysis on the mRNA view.
Here we are going to use the absolute value of the weight (or loading) as feature statistic, the mean difference between the features in the gene set and the features outside the gene set as a feature set statistic and the correlation-adjusted parametric statistical test.
```{r}
# The input format for feature sets has to be a binary matrix where the rows represent feature sets and the columns are features where a value of 1 indicates that the feature belongs to the corresponding feature set.

# Load reactome annotations
reactome <- readRDS("/Users/ricard/data/reactome/v59/homo_sapiens/out/human_reactome.rds")

# Run Feature Set Enrichment Analysis
p.values <- FeatureSetEnrichmentAnalysis(
  model,
  view = "mRNA", 
  factors = "all", 
  feature.sets = reactome,
  local.statistic = "loading",
  transformation = "abs.value",
  global.statistic = "mean.diff",
  statistical.test = "cor.adj.parametric",
  nperm=NA, 
  min.size=10,
  cores=1
)

# Rename pathways
# tmp <- read.table("/Users/ricard/data/reactome/v59/homo_sapiens/AllPathways.txt", header=F, quote="", sep="\t", stringsAsFactors=F)[,c(1,2)]
# reactome_meta <- tmp[,2]; names(reactome_meta) <- tmp[,1]
# rownames(p.values) <- stringr::str_replace_all(rownames(gsea.out_reactome), reactome_meta)

# Correct for multiple testing
p.values_adj <- apply(p.values, 2,function(lfw) p.adjust(lfw, method = "BH"))


# Visualise the results 
# all factors
Heatmap_FeatureSetEnrichmentAnalysis(p.values_adj, threshold=0.1)

# individual factors
LinePlot_FeatureSetEnrichmentAnalysis(p.values_adj, view, factor, threshold=0.1, max.pathways=25)
```

# Step 4: Impute Missing observations
```{r}
imputeMissing(object)
```


<!-- # Cluster samples -->
<!-- ```{r} -->
<!-- clusterMOFA(object, factors) -->
<!-- ``` -->

